---
title: "Lab 6: extra practice"
author: "Kristi Gdanetz MacCready"
date: "02/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMBio201/Lab6/")
```

# Load packages
```{r Load packages, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Import data
Load the curated data created during Lab 6 if it is not already in your global environment
```{r}
avg_qc <- read_delim(file = "curated_data/avg_qc.txt", 
                      delim = "\t", na = "NA", col_names = TRUE)

bio2 <- read_delim(file = "curated_data/biographical_data.txt", 
                   delim = "\t", na = "NA", col_names = TRUE)

join5 <- read_delim(file = "curated_data/joined_results.txt", 
                    delim = "\t", na = "NA", col_names = TRUE)
```


# Pratice 1
Join the biographical and join5 data frames. Use this new data frame to answer the following questions:

* (Q1) is the proportion of male and female participants consuming starch different?
* (Q2) is the proportion of students with a positive response (this can be with one of more measures, SCFA, pH, etc.) to any potato starch different by sex? 
* (Q3) Do you observe the same pattern when comparing types (LOODAT or BRMPS) of potato starches?
* (Q4) Do you observe the same pattern when comparing frequency (1x or 2x) of starch consumed?
* (Q5) is the proportion of students with a positive response to potato starch different by race/ethnicity? 
* (Q6) Do you observe the same pattern when comparing types (LOODAT or BRMPS) of potato starches?
* (Q7) Do you observe the same pattern when comparing frequency (1x or 2x) of starch consumed?

```{r}
# join here 
```

### Q1
```{r}

```

### Q2
```{r}

```

### Q3
```{r}

```

### Q4
```{r}

```

### Q5
```{r}

```

### Q6
```{r}

```

### Q7
```{r}

```


# Practice 2

###  Filter and plot
Repeat the spread + gather process from Lab 6 to reduce the number of individuals who have paired measurements for pH and Bristol score. Calculate delta values for each of these metrics and plot them. 
```{r}

```

### Q1
Check assumptions, select and conduct the test you would need to use to identify if pH changed when participants consumed BRMPS. Hint: do not use the delta values here. 
```{r}

```

### Q3
Check assumptions, select and conduct the test you would need to use to identify under which supplement there was the greatest change in pH?
```{r}

```


-----
end