---
title: "Lab 6: extra practice"
author: "Kristi Gdanetz MacCready"
date: "02/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Box Sync/Schmidt_Lab/Bio201_teaching/W20_Protocols/Lab6/")
```

# Load packages
```{r Load packages, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Import data
Load the curated data created during Lab 6 if it is not already in your global environment
```{r}
avg_qc <- read_delim(file = "curated_data/avg_qc.txt", 
                      delim = "\t", na = "NA", col_names = TRUE)

bio2 <- read_delim(file = "curated_data/biographical_data.txt", 
                   delim = "\t", na = "NA", col_names = TRUE)

join5 <- read_delim(file = "curated_data/joined_results.txt", 
                    delim = "\t", na = "NA", col_names = TRUE)
```


# Pratice 1
Join the biographical and join5 data frames. Use this new data frame to answer the following questions:

* (Q1) is the proportion of male and female participants consuming starch different?
* (Q2) is the proportion of students with a positive response (this can be with one of more measures, SCFA, pH, etc.) to any potato starch different by sex? 
* (Q3) Do you observe the same pattern when comparing types (LOODAT or BRMPS) of potato starches?
* (Q4) Do you observe the same pattern when comparing frequency (1x or 2x) of starch consumed?
* (Q5) is the proportion of students with a positive response to potato starch different by race/ethnicity? 
* (Q6) Do you observe the same pattern when comparing types (LOODAT or BRMPS) of potato starches?
* (Q7) Do you observe the same pattern when comparing frequency (1x or 2x) of starch consumed?

```{r}
# join here 
joined <- inner_join(bio2, join5, by = "participant_id")
```

### Q1
```{r}

```

### Q2
```{r}

```

### Q3
```{r}

```

### Q4
```{r}

```

### Q5
```{r}

```

### Q6
```{r}

```

### Q7
```{r}

```


# Practice 2

###  Filter and plot
Repeat the spread + gather process from Lab 6 to reduce the number of individuals who have paired measurements for pH and Bristol score. Calculate delta values for each of these metrics and plot them. 
```{r}
# calculate detla pH
pH_wide <- avg_qc %>%
  select(participant_id, study_week, frequency, supplement_consumed, pH_mean) %>% 
  filter(study_week == "week1" | study_week == "week3" ) %>% 
  spread(key = "study_week", value = "pH_mean") %>%
  mutate(week1 = as.numeric(week1),
         week3 = as.numeric(week3),
         delta_ph = week3 - week1) %>%
  drop_na(delta_ph) %>%
  select(-week1, -week3)
pH_wide
```
```{r}
# calculate detla Bristol
bss_wide <- avg_qc %>%
  select(participant_id, study_week, frequency, supplement_consumed, bristol_mean) %>% 
  filter(study_week == "week1" | study_week == "week3" ) %>% 
  spread(key = "study_week", value = "bristol_mean") %>%
  mutate(week1 = as.numeric(week1),
         week3 = as.numeric(week3),
         delta_bss = week3 - week1) %>%
  drop_na(delta_bss) %>%
  select(-week1, -week3)
bss_wide
```
```{r}
# join data frames with deltas and plot 
ph_bss <- full_join(pH_wide, bss_wide, 
                    by = c("participant_id", "frequency", "supplement_consumed"))

# pH plot
ph_bss %>%
  ggplot(aes(x = supplement_consumed, y = delta_ph, 
             color = supplement_consumed)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_violin() +
  geom_jitter(width = 0.25) + 
  xlab(NULL) + 
  scale_y_continuous(name = "pH (wk3 - wk1)",
                     breaks = c(-2, -1, 0, 1, 2)) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# BSS plot
ph_bss %>%
  ggplot(aes(x = supplement_consumed, y = delta_bss, 
             color = supplement_consumed)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_violin() +
  geom_jitter(width = 0.25) + 
  xlab(NULL) + 
  scale_y_continuous(name = "Bristol Score (wk3 - wk1)",
                     breaks = c(-3, -2, -1, 0, 1, 2, 3)) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Q1
Check assumptions, select and conduct the test you would need to use to identify if pH changed when participants consumed BRMPS. Hint: do not use the delta values here. 
```{r}
pH_wide <- avg_qc %>%
  select(participant_id, study_week, frequency, supplement_consumed, pH_mean) %>% 
  filter(study_week == "week1" | study_week == "week3" ) %>% 
  spread(key = "study_week", value = "pH_mean") %>%
  mutate(week1 = as.numeric(week1),
         week3 = as.numeric(week3)) %>%
  drop_na() %>%
  filter(supplement_consumed == "BRMPS")

# check sample size 
pH_wide %>%
  summarise(N = n())
# check variances
var.test(x = pH_wide$week1, y = pH_wide$week3, paired = TRUE, alternative = "two.sided") #p-value = 0.1696

# t-test
t.test(x = pH_wide$week1, y = pH_wide$week3,
       paired = TRUE, alternative = "two.sided", var.equal = TRUE) #0.01562
```

### Q3
Check assumptions, select and conduct the test you would need to use to identify under which supplement there was the greatest change in pH?
```{r}
# sample size 
ph_bss %>%
  select(-delta_bss) %>%
  drop_na() %>% 
  group_by(supplement_consumed) %>%
  summarise(Total = n())

# variance check = formula of the form lhs ~ rhs where lhs gives the data values and rhs the corresponding groups.
ph_bss %>%
  select(-delta_bss) %>%
  drop_na() %>% 
  bartlett.test(., formula = delta_ph ~ supplement_consumed)

kruskal.test(delta_ph ~ supplement_consumed, data = ph_bss) #formula = response ~ group 
```


-----
end